\documentclass[twocolumn]{svjour3} 

\usepackage[square,sort,comma,numbers]{natbib}
\usepackage{graphicx}
\usepackage{amssymb,amstext}
\usepackage{url}

\journalname{Journal on Multimodal User Interfaces}

\begin{document}

\title{Emergence of Turn-Taking Behaviors in User-Agent Interactions}

\author{Mathieu J\'egou  \and
        Pierre Chevaillier
}

\institute{M. J\'egou \at
              B-Com, France \\
              Tel.: +33-XXX-XX-XX-XX\\
              \email{mathieu.jegou@gmail.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           P. Chevaillier \at
              Lab-STICC, ENIB, F-29280 Plouzan\'e, France
}

\date{Received: date / Accepted: date}

\maketitle 

\begin{abstract}
We propose a computational model that endows conversational agents with the capability to coordinate their speaking turns in the context of mixed-initiative dialogs.
In human conversations, participants are continuously adjusting their verbal and non verbal productions to adapt their behavior for maintaining effective the coordination. Turn-taking management needs for the agents to make a decision whether to keep on speaking (or listening) or not. In our model, the decision-making is a continuous process based on the intrinsic current goal of the agent with respect to turn-taking, namely its motivation to keep its current role (speaker or listener) or to switch to the opposite role and on its perception of the intentions of its partner. Concurrently, the agent is also producing signals indicating its willingness to keep or to leave its current role. Because all the participants in the interaction behave accordingly to this principle, the actual coordination of speaking turns is an emergent property. It results from the continuous sensory-motor coupling of the interactants. Our model is based on two models from the cognitive psychology: the drift-diffusion model and the theory of the behavioral dynamics. After showing on simulations how our model makes the coordination to emerge from the interaction, we propose a SAIBA-Compliant architecture, named BeAware, created to support the implementation of our model. Finally, based on the implementation of our model, we investigated how the agent's turn-taking strategy may impact the user's experience and the effectiveness of the coordination.

 
%Insert your abstract here. Include keywords, PACS and mathematical subject classification numbers as needed.
\keywords{Turn-taking \and User-Agent interaction \and Embodied-conversational-agent, prosody}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\input{intro_gene.tex}

\input{related.tex}

\input{theoretical_model.tex}

\input{model_properties.tex}

\input{implementation.tex}

\input{ua_inter.tex}

\input{conclusion.tex}

\bibliographystyle{spmpsci}

\bibliography{jmui}

\end{document}